---
title: "Practical Machine Learning Project"
author: "Joel Chavez Gomez"
date: "23/1/2022"
output: html_document
theme: readable
---
## Data 
The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 

## What you should submit
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

## Assignment
Libraries are loaded, data sets files are downloaded and read.
```{r setup, messaage = FALSE, warning = FALSE, cache = TRUE}
library(tidyverse); library(caret); library(kableExtra);
Train.url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
Test.url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
if(!file.exists("DataSets/pml-training.csv")){
download.file(Train.url, destfile = "DataSets/pml-training.csv")
}
if(!file.exists("DataSets/pml-testing.csv")){
download.file(Test.url, "DataSets/pml-testing.csv")
}
training <- read.csv("./DataSets/pml-training.csv", header = TRUE)
testing <- read.csv("./DataSets/pml-testing.csv", header = TRUE)
```

## Exploratory Data Analysis?

Adjusting the database
```{r eda, warning = FALSE, message = FALSE, cache = TRUE}
dim(training)
names(training)
train_data <- training[-c(1:7)]
train_data <- sapply(train_data[, -c(153)], as.numeric)
train_data <- as.data.frame(train_data)
#train_data$classe <- as.factor(training$classe)
```

Data Preprocessing

```{r prepro, cache = TRUE}
# Removing features with near zero variance
train_nzv <- nearZeroVar(train_data)
train_data <- train_data[, -train_nzv]

# Removing highly correlated features
M <- abs(cor(train_data))
M[!lower.tri(M)] <- 0
high_cor_features <- which(M > 0.8, arr.ind = TRUE)
train_data.new <- train_data[, !apply(M, 2, function(x) any(abs(x) > 0.8, na.rm = TRUE))]

# Removing features with NA values
train_data.new <- train_data.new[lapply(train_data.new, 
                                        function(x) sum(is.na(x))/length(x) ) < 0.9 ]

# Adding outcome var
train_data.new$classe <- training$classe
train_data.new$classe <- as.factor(train_data.new$classe)
```

```{r feat-plot}
fitControl <- trainControl(method = "cv", number = 3)
```

```{r models}
gbmFit1 <- train(classe ~ ., data = train_data.new,
                 method = "gbm", 
                 trControl = fitControl, 
                 verbose = FALSE)
rfFit <- train(classe ~ ., data = train_data.new, 
               trControl = fitControl,
               method = "rf")
```

```{r pred}
rf_pred <- predict(rfFit, testing)
```